Parsing Postfix log files

### 2007-04-12 doc/logparser.tex
Added "Database as API" section because it seemed warranted when trying to expand the motivation section. Added example SQL queries to show what can be done with the data.  Should probably be moved to an appendix at some stage. Fixed minor typographical issues picked up by chktex. Added more place-holder XXX items.

XXX WHAT IS THE CONCLUSION??? XXX

XXX FIGURE OUT WHEN TO USE MAIL AND WHEN TO USE CONNECTION; POSSIBLY STATE
THAT THEY'RE MORE OR LESS EQUIVALENT  XXX

@@

@@

Section~\ref{conclusion} describes the results of the project to date.  XXX
IMPROVE THIS.

@@

This paper and the program it describes are part of a larger project to
optimise a server's Postfix restrictions, generate statistics and graphs,
and provide a platform on which new restrictions can be trialled and
evaluated to see if they are worth using in the fight against spam.  The
program parses Postfix logs and populates a database with the data gleaned
from those logs, providing a consistent and simple view of the logs which
future tools can utilise.

XXX EXPLAIN WHY THIS IS USEFUL AND IMPORTANT XXX

XXX ADD A SECTION EXPLAINING THE IMMEDIATE BENEFITS \\
\texttt{SELECT name, description, restriction\_name, rule\_order \\
    FROM rules \\
    ORDER BY rule\_order;}\\
Tells you which of your restrictions are most effective.

XXX COMPARE TO PREVIOUS WORK XXX

@@

XXX ADD INFORMATION ABOUT OTHER POSTFIX LOG PARSERS HERE XXX

@@

        \_\_COMMAND\_\_ is an SMTP command, e.g. MAIL FROM; hopefully the
        others are reasonably self explanatory.

@@

@@

    \item [result]  XXX RESULT NEEDS TO BE RENAMED XXX This is the action
        Postfix must have taken to generate this line, with two exceptions:

@@

        XXX RERUN OVER MORE LOG FILES XXX

@@

XXX CARL SUGGESTS EXPANDING ON THIS SOME MORE, BUT I DON'T KNOW HOW\@.
PERHAPS WHEN THE ALGORITHM IS PROPERLY EXPLAINED IT'LL BE EASIER\@?  MAKE
REFERENCES TO THE ALGORITHM, EXPLAIN MORE CLEARLY HOW THE FIELDS ARE USED,
HOW THE RULE MATCHES, ETC.  XXX

@@

name                & Unknown sender domain                             \\
description         & We do not accept mail from unknown domains        \\
program             & postfix/smtpd                                     \\
restriction\_name   & reject\_unknown\_sender\_domain                   \\
regex               & \verb!^<(__SENDER__)>: Sender address rejected: ! \\
                    & \verb!  Domain not found; from=<\1> !             \\
                    & \verb!  to=<(__RECIPIENT__)> proto=E?SMTP !       \\
                    & \verb!  helo=<(__HELO__)>$!                       \\
result\_cols        & recipient = 2; sender = 1                         \\
connection\_cols    & helo = 3                                          \\
result\_data        &                                                   \\
connection\_data    &                                                   \\
result              & REJECTED                                          \\
action              & SAVE\_BY\_PID                                     \\
queueid             & 0                                                 \\
rule\_order         & 0                                                 \\
priority            & 0                                                 \\

@@

While the rules are more than double (treble with 2.3.x rules? XXX ADD
2.3.X RULES AND CHECK XXX) the size of
the algorithm, the rules are quite simple and each rule is completely
independent of its fellows.  The algorithm is significantly more
complicated and highly internally interdependent.

@@

    \item Multiple independent mails may be delivered during one
        connection: this requires cloning the current data as soon as a
        mail is accepted, so that subsequent mails won't trample over each
        other's data.  This must be done every time a mail is accepted, as
        it's impossible to tell in advance which connections will accept
        multiple mails.  It is quite easy to overlook this complication
        because only a small minority of connections accept more than one
        mail. Happily once the mail has been accepted log entries won't be
        correlated by pid for that mail any more (its queueid will be used
        instead), so there isn't any ambiguity about which mail a given log
        line belongs to.  The original connection will be discarded unsaved
        when the client disconnects.  One unsolved difficulty is
        distinguishing between different groups of rejections, e.g. when
        dealing with the following sequence:

@@

        Process scheduling can have a still more confusing effect: quite
        often the child mail will be created, delivered and entirely
        finished with \textbf{before} the parent logs the creation line!
        Thus mails marked as faked cannot be entered into the database when
        their final log line is parsed; instead they must be marked as
        database ready and subsequently entered by the parent mail once it
        has been identified.  Quite apart from identifying mail injected in
        an unknown fashion, or bugs in the parser, faked mails need to be
        marked as such because they lack some data present in the parent
        mail, and must copy that data from their parent before being
        entered in the database.  XXX EITHER REMOVE THE LAST LINE OR EXPAND
        ON IT SOME MORE.  FIGURE OUT WHY MORE DATA ISN'T REQUIRED.

@@

XXX USE A DISPATCH TABLE TO MAKE THIS ACCURATE XXX

@@

@@

XXX ADD A LIST IN LOAD\_RULES() SPECIFYING THE ALLOWED ACTIONS XXX

@@

XXX ADD RESOURCES ONCE THE PAPER IS REASONABLY COMPLETE AND THEN REFERENCE
THEM FROM WITHIN THE PAPER XXX

>>>>

XXX WHAT IS THE CONCLUSION???

XXX FIGURE OUT WHEN TO USE MAIL AND WHEN TO USE CONNECTION\@; POSSIBLY STATE
THAT THEY'RE MORE OR LESS EQUIVALENT

@@

\label{introduction}

@@

Section~\ref{conclusion} describes the results of the project to date.  XXX
IMPROVE THIS\@.

@@

This paper and the program it describes are part of a larger project to
optimise a server's Postfix restrictions, generate statistics and graphs,
and provide a platform on which new restrictions can be trialled and
evaluated to see if they are worth using in the fight against spam.  The
program parses Postfix logs and populates a database with the data gleaned
from those logs, providing a consistent and simple view of the logs which
future tools can utilise.  The gathered data can then be used to optimise
current anti-spam measures, provide a baseline to test new anti-spam
measures against, or to produce statistics showing how effective those
measures are.\footnote{Section~\ref{introduction} lists some more example
uses.}

A short example of the optimisation possible using data from the database
is optimal ordering of Postfix restrictions:

\begin{verbatim}
SELECT name, description, restriction_name, rule_order
    FROM rules
    WHERE postfix_action = 'REJECTED'
    ORDER BY rule_order DESC;
\end{verbatim}

If the database supports sub-selects percentages can be
obtained:\footnote{This works for SQLite3, where || is the concatenation
operator.  If the database containing the data does not support
concatenation, simply remove ``|| '\%\''' from the query.}

XXX STOP THIS FROM BEING SPLIT ACROSS PAGES IF POSSIBLE

\begin{verbatim}
SELECT name, description, restriction_name, rule_order,
        (rule_order * 100.0 /
            (SELECT SUM(rule_order)
                FROM rules
                WHERE postfix_action = 'REJECTED'
            )
        ) || '%' AS percentage
    FROM rules
    WHERE postfix_action = 'REJECTED'
    ORDER BY rule_order DESC;
\end{verbatim}

Another example is determining which restrictions are not effective:

\begin{verbatim}
SELECT name, description, restriction_name, rule_order,
        (rule_order * 100.0 /
            (SELECT SUM(rule_order)
                FROM rules
                WHERE postfix_action = 'REJECTED'
            )
        ) || '%' AS percentage
    FROM rules
    WHERE postfix_action = 'REJECTED'
        AND rule_order < 100
    ORDER BY rule_order ASC;
\end{verbatim}

XXX COMPARE TO OTHER PARSERS

\subsection{Database as Application Programming Interface}

The database populated by this program provides a simple interface to
Postfix logs.  Although the interface is a database schema, it is in effect
quite similar to the API provided by a library: it insulates both user and
provider of the API from changes in the implementation of the other.  The
algorithm implemented by the program can be improved; support can be added
for earlier or later releases of Postfix; bugs can be fixed or limitations
removed from the program.  Statistics and/or graphs can be generated from
the database; new restrictions can be tested and the results inspected;
trends in the fight against spam can emerge from historical data saved in
the database.  Using a database simplifies writing programs which need to
interact with the data in several ways:

    \item Libraries exist for the majority of programming languages
        allowing access to databases, whereas if this program was written
        as a library a new interface layer would need to be written for
        every programming language used with the API\@.

    \item Databases provide complex querying and sorting functionality to
        the user without requiring large amounts of programming.  All
        databases provide one or more programs, of variable complexity and
        sophistication, which can be used for ad hoc queries with minimal
        investment of time.

    \item Databases are easily extensible, e.g.:

            \item Other tables can be added to the database.

            \item New columns can be added to the tables used by the
                program, with sufficient DEFAULT clauses or a clever
                TRIGGER or two.

            \item A VIEW gives a custom arrangement of data with very
                little effort.

            \item If the database supports it, access can be granted on a
                fine-grained basis so that the finance department can
                produce invoices, the helpdesk can run limited queries as
                part of dealing with support calls, and the administrators
                have full access to the data.

            \item Triggers can be written to perform actions when certain
                events occur.  In pseudo-SQL\@:

\begin{verbatim}
CREATE TRIGGER ON INSERT INTO results
    WHERE sender = 'boss@example.com'
        AND postfix_action = 'REJECTED'
    SEND PANIC EMAIL TO 'postmaster@example.com';
\end{verbatim}

@@

XXX ADD INFORMATION ABOUT OTHER POSTFIX LOG PARSERS HERE

@@

        \_\_COMMAND\_\_ is an SMTP command, e.g. MAIL FROM\@; hopefully the
        others are reasonably self explanatory.

@@

        XXX RERUN OVER MORE LOG FILES

@@

    \item [postfix\_action] This is the action Postfix must have taken to
        generate this line, with two exceptions:

@@

        XXX RERUN OVER MORE LOG FILES

@@

XXX CARL SUGGESTS EXPANDING ON THIS SOME MORE, BUT I DON'T KNOW HOW\@.
PERHAPS WHEN THE ALGORITHM IS PROPERLY EXPLAINED IT'LL BE EASIER\@?  MAKE
REFERENCES TO THE ALGORITHM, EXPLAIN MORE CLEARLY HOW THE FIELDS ARE USED,
HOW THE RULE MATCHES, ETC\@.

@@

name                & Unknown sender domain                             \\
description         & We do not accept mail from unknown domains        \\
program             & postfix/smtpd                                     \\
restriction\_name   & reject\_unknown\_sender\_domain                   \\
regex               & \verb!^<(__SENDER__)>: Sender address rejected: ! \\
                    & \verb!  Domain not found; from=<\1> !             \\
                    & \verb!  to=<(__RECIPIENT__)> proto=E?SMTP !       \\
                    & \verb!  helo=<(__HELO__)>$!                       \\
result\_cols        & recipient = 2; sender = 1                         \\
connection\_cols    & helo = 3                                          \\
result\_data        &                                                   \\
connection\_data    &                                                   \\
postfix\_action     & REJECTED                                          \\
action              & SAVE\_BY\_PID                                     \\
queueid             & 0                                                 \\
rule\_order         & 0                                                 \\
priority            & 0                                                 \\

@@

While the rules have more lines of code than the algorithm, the rules are
quite simple and each rule is completely independent of its fellows.  The
algorithm is significantly more complicated and highly internally
interdependent.

@@

    \item Multiple independent mails may be delivered during one
        connection: this requires cloning the current data as soon as a
        mail is accepted, so that subsequent mails won't trample over each
        other's data.  This must be done every time a mail is accepted, as
        it's impossible to tell in advance which connections will accept
        multiple mails.  It is quite easy to overlook this complication
        because only a small minority of connections accept more than one
        mail. Happily once the mail has been accepted log entries won't be
        correlated by pid for that mail any more (its queueid will be used
        instead), so there isn't any ambiguity about which mail a given log
        line belongs to.  The original connection will be discarded unsaved
        when the client disconnects.  One unsolved difficulty is
        distinguishing between different groups of rejections, e.g.\ when
        dealing with the following sequence:

@@

        Process scheduling can have a still more confusing effect: quite
        often the child mail will be created, delivered and entirely
        finished with \textbf{before} the parent logs the creation line!
        Thus mails marked as faked cannot be entered into the database when
        their final log line is parsed; instead they must be marked as
        database ready and subsequently entered by the parent mail once it
        has been identified.  Quite apart from identifying mail injected in
        an unknown fashion, or bugs in the parser, faked mails need to be
        marked as such because they lack some data present in the parent
        mail, and must copy that data from their parent before being
        entered in the database.  XXX EITHER REMOVE THE LAST LINE OR EXPAND
        ON IT SOME MORE\@.  FIGURE OUT WHY MORE DATA ISN'T REQUIRED\@.

@@

@@

XXX EXPAND THIS

@@

XXX USE THE SAME DOCUMENTATION HERE AS IN Parser.pm

@@

XXX ADD RESOURCES ONCE THE PAPER IS REASONABLY COMPLETE AND THEN REFERENCE
THEM FROM WITHIN THE PAPER
